---
title: "Curriculum Vitae"
author: Dr Matt Ashby
date: last-modified
address: UCL Jill Dando Institute of Security and Crime Science, 35 Tavistock Square, London WC1H 9EZ
phone: +44 (0) 20 3108 3890
email: matthew.ashby@ucl.ac.uk
web: mattashby.com
format: 
  quarto-cv-pdf: 
    include-in-header: 
      - text: |
          \usepackage[none]{hyphenat}
          \renewcommand{\descriptionlabel}[1]{\hspace{\labelsep}\textrm{#1}}
execute: 
  echo: false
---


```{r setup}
#| include: false

library(gt)
library(rorcid)
library(tidyverse)

my_orcid <- "0000-0003-4201-9239"
```



## Employment

```{r employment}
employment <- orcid_employments(my_orcid) |> 
	pluck(my_orcid, "affiliation-group", "summaries") |> 
  bind_rows() |> 
  rename_with(\(x) str_remove(x, "^employment-summary.")) |> 
  janitor::clean_names() |> 
  mutate(
    date_start = format(
      make_date(start_date_year_value, start_date_month_value), 
      "%b %Y"
    ),
    date_end = if_else(
      !is.na(end_date_year_value),
      format(make_date(end_date_year_value, end_date_month_value), "%b %Y"),
      "present"
    ),
    when = str_glue("{date_start}--{date_end}"),
    what = str_glue("**{role_title}**, {department_name}, {organization_name}"),
    output = str_glue("{when}\n: {what}")
  ) |> 
  pull("output")
```

`r str_flatten(employment, collapse = "\n\n")`



## Education

```{r education}
education <- orcid_educations(my_orcid) |> 
	pluck(my_orcid, "affiliation-group", "summaries") |> 
  bind_rows() |> 
  rename_with(\(x) str_remove(x, "^education-summary.")) |> 
  janitor::clean_names() |> 
  mutate(
    date_start = format(
      make_date(start_date_year_value, start_date_month_value), 
      "%b %Y"
    ),
    date_end = if_else(
      !is.na(end_date_year_value),
      format(make_date(end_date_year_value, end_date_month_value), "%b %Y"),
      "present"
    ),
    when = str_glue("{date_start}--{date_end}"),
    what = str_glue("**{role_title}**, {organization_name}"),
    output = str_glue("{when}\n: {what}")
  ) |> 
  pull("output")
```

`r str_flatten(education, collapse = "\n\n")`



## Professional qualifications

```{r qualifications}
qualifications <- orcid_qualifications(my_orcid) |> 
	pluck(my_orcid, "affiliation-group", "summaries") |> 
	bind_rows() |> 
  rename_with(\(x) str_remove(x, "^qualification-summary.")) |> 
  janitor::clean_names() |> 
  mutate(
    output = str_glue(
      "{start_date_year_value}\n: **{role_title}**, {organization_name}"
    )
  ) |> 
  pull("output")
```

`r str_flatten(qualifications, collapse = "\n\n")`



## Publications

ORCID: [`r my_orcid`](https://orcid.org/`r my_orcid`)

```{r publications}
publications <- scholar::get_publications("dwcesrQAAAAJ") |> 
  filter(!is.na(year)) |> 
  arrange(desc(year)) |> 
  mutate(
    authors = str_split(author, ", "),
    authors = map(
      authors,
      function(x) {
        case_when(
          str_detect(x, "Ashby") ~ str_glue("**M Ashby**"), 
          x == "..." ~ "_et al._", 
          TRUE ~ x
        )
      }
    ),
    authors = map_chr(authors, \(x) str_flatten(x, collapse = ", ")),
    publication = str_glue("_{journal}_ {number}"),
    output = str_glue("{authors} ({year}).\n: [_{str_to_title(title)}_](https://scholar.google.com/citations?view_op=view_citation&hl=en&user=dwcesrQAAAAJ&citation_for_view=dwcesrQAAAAJ:{pubid}). {publication}.")
  ) |> 
  pull("output")
```

`r str_flatten(publications, collapse = "\n\n")`


## Funding

Funding awarded, typically jointly with colleagues.

```{r funding}
funding <- tribble(
	~type, ~title, ~start_date, ~end_date, ~value, ~funder, ~role,
	"Knowledge transfer partnership", "Problem Solving for Neighbourhood Policing", 2012, 2013, "£16,000", "Engineering and Physical Sciences Research Council", "co-investigator",
	"Vice Chancellor's PhD studentship", "Crime reduction and public reassurance in a diverse society", 2017, 2020, "£60,000", "Nottingham Trent University", "supervisor",
	"Vice Chancellor's PhD studentship", "Building evidence to support police initial response to incidents", 2018, 2021, "£60,000", "Nottingham Trent University", "supervisor",
	"Research project", "Deciphering and disrupting the social, spatial and temporal systems behind transnational human trafficking: a data-science approach", 2018, 2023, "£365,000", "Economic and Social Research Council", "co-investigator",
	"Research project", "Problem-Oriented Policing in England and Wales", 2019, 2020, "£246,000", "Police Transformation Fund", "co-investigator",
	"Research project", "Building blocks for knife crime", 2020, 2021, "£122,000", "College of Policing", "co-investigator",
	"Knowledge transfer", "Implementing problem-oriented and evidence-based responses for reducing violent crime in London", 2021, 2022, "£264,000", "Metropolitan Police Service", "co-investigator",
	"Research project", "Benchmarking crime in London against other global cities", 2022, NA, "£10,000", "Metropolitan Police Service", "Principal investigator",
	"Research project", "National homicide and near-miss homicide problem profile", 2023, 2024, "£63,000", "College of Policing", "Principal investigator"
) |> 
	arrange(desc(start_date)) |> 
  mutate(
    when = if_else(
      !is.na(end_date),
      str_glue("{start_date}--{end_date}"),
      str_glue("{start_date}")
    ),
    what = str_glue(
      "{type}: **{title}**. {str_to_sentence(role)}, {value} funding from the ",
      "{funder}."
    ),
    output = str_glue("{when}\n: {what}")
  ) |> 
  pull("output")
```

`r str_flatten(funding, collapse = "\n\n")`



## Software

I maintain the following packages for the R statistical language:

 * `sfhotspot`: a package for analysing geographic hotspots of events such as crimes.
 * `crimemappingdata`: a package providing open-source datasets from around the world that are suitable for teaching crime mapping.



## Professional service

```{r service}
service <- tribble(
  ~name, ~start_date, ~end_date, ~desc, ~url,
  "CrimPapers journal alert service", 2017, NA, "I maintain the CrimPapers alert service for new articles in 100+ criminology and criminal justice journals", "https://lesscrime.info/crimpapers/",
  "Crime Open Database", 2018, NA, "I maintain the Crime Open Database, a dataset of harmonised open crime data for 18 large cities in the United States, as well as the associated `crimedata` R package", "http://osf.io/zyaqn/",
  "Associate editor of the journal _Crime Science_", 2019, 2024, NA, NA,
  "Judge at the London Problem Solving Awards", 2019, 2021, "I was a judge for the joint problem-solving awards run by the three main police forces in London", NA,
  "Sift judge for the Tilley Awards", 2023, NA, "I am a judge for the initial round of the Tilley Awards for problem solving run by UK police forces", NA,
  "Editorial board of the journal _Criminology_", 2024, NA, NA, NA
) |> 
  arrange(desc(start_date)) |> 
  mutate(
    when = if_else(
      !is.na(end_date),
      str_glue("{start_date}--{end_date}"),
      str_glue("{start_date}--present")
    ),
    desc = if_else(!is.na(desc), str_glue("{desc}."), ""),
    what = str_glue("**{name}**. {desc}"),
    output = str_glue("{when}\n: {what}")
  ) |> 
  pull("output")
```

`r str_flatten(service, collapse = "\n\n")`


